Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-15 Paper/Paper.tex
added some drawings

\section{Related Work}

@@

@@

In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. For example, Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense and classify human postures \cite{Wimmer,Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

@@

The 3D accelerometer sensor is being sampled at $100Hz$, resulting in 10ms equidistant measurements. For efficiency reasons, the sensor data is run-length encoded before being stored locally to the microSD card.

The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor, which is described in detail in the next section.

\subsection{Capacitive Proximity Sensor}
power drain: 1mA when active
image : photo
image : board layout and schematics !

The capacitive proximity sensor for the wristband performs measurements in loading mode. The sensing circuit schematic is shown in Figure \ref{}. It is based on a timer that controls the charging and discharging cycles of the virtual capacitor. The timer toggles from charge to discharge at the time when a threshold voltage at the capacitor is reached. This results in an astable operation with succeeding charge/discharge cycles. When the capacitance of the virtual capacitor increases, the charging time will also increase and vice-versa. Therefore, the capacitance is inversely proportional to the number of charging cycles in a given time span. In order to guard the sensor from measuring the capacitance to underlying objects, a shield is placed directly underneath the measuring electrode. The shield is driven with the same potential as the sensing electrode, such that the the capacitance between the two electrodes is neglectable. Using this shielding method, the measured capacitance will only be slightly affected by the grounded underlying arm.

The proximity sensor board generates a clock signal with varying frequency depending on the proximity to body parts, objects and their characteristics. The HedgeHog therefore needs to measure the incoming frequency, which is implemented by counting the signal's edges over the gate time of approximately 9.5ms.

pseudocode:
    sample the accelerometer sensor

start frequency counter

go to sleep for 9.5 ms

wake-up (timer interrupt)

stop frequency counter

read frequency count

\section{Experiment}

@@

The recorded test set contains the following activities: opening door, sitting on a couch, lying on a couch, grasping into a locker, putting something from a shelf, sitting at the table, preparing a bread with marmalade, eating the bread, pouring water, drinking the water and sleeping. Some of those activities are very hard to recognize without context information, just using wrist-worn accelerometer data. For example, sitting at the table and sitting on a couch are very similar activities. The data of a capacitive proximity sensor can extend this data basis by providing that is related to the material an object is made of.

>>>>

\section{Related work}

@@

\begin{figure}[htbp]
	\centering
		\includegraphics[width=1.00\textwidth]{Images/modes.png}
	\caption{Capacitive proximity sensing modes \cite{Smith1996}}
	\label{fig:modes}
\end{figure}

@@

In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. For example, Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

@@

The 3D accelerometer sensor is being sampled at $100Hz$, resulting in 10ms equidistant measurements. For efficiency reasons, the sensor data is run-length encoded before being stored locally to the microSD card. The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor, which is described in detail in the next section.

\subsection{Capacitive proximity sensor}
power drain: 1mA when active
image : photo
image : board layout and schematics !

The capacitive proximity sensor for the wristband performs measurements in loading mode. The sensing circuit schematic is shown in Figure \ref{}. It is based on a timer that controls the charging and discharging cycles of the virtual capacitor that is built by the wristband electrode and the surrounding environment. The timer toggles from charge to discharge at the time when a threshold voltage at the capacitor is reached. This results in an astable operation with succeeding charge/discharge cycles. When the capacitance of the virtual capacitor increases, the charging time will also increase and vice-versa. Therefore, the capacitance is inversely proportional to the number of charging cycles in a given time span. In order to guard the sensor from measuring the capacitance to underlying objects, a shield is placed directly underneath the measuring electrode. The shield is driven with the same potential as the sensing electrode, such that the the capacitance between the two electrodes is neglectable. Using this shielding method, the measured capacitance will only be slightly affected by the grounded underlying arm.

The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms.

\begin{figure}[htbp]
	\centering
		\includegraphics[trim=1cm 1cm 1cm 1cm,width=1.00\textwidth]{Images/pseudocode.pdf}
	\caption{Measurement procedure}
	\label{fig:pseudocode}
\end{figure}
\section{Experiment}

@@

The recorded test set contains the following activities: opening door, sitting on a couch, lying on a couch, grasping into a locker, putting something from a shelf, sitting at the table, preparing a bread with marmalade, eating the bread, pouring water, drinking the water and sleeping. Some of those activities are very hard to recognize without any further contextual information, when the data is limited to a single modality like a 3D accelerometer. For example, sitting at the table and sitting on a couch are very similar activities. The data of a capacitive proximity sensor can extend this data basis by providing that is related to the material an object is made of.
